---
title: "Five Recommended Flights"
output: html_document
---
# Summary of Findings
- The top five flights are DFW-OGG, EWR-SFO, HNL-SFO, HNL-IAH, ATL-HNL. These would be my recommendations as they would provide the most profit in the shortest amount of time. 
- The top flights for profitability are different than the busiest flights. 
- Key metrics to track: profitability, average delays, ticket fares, break even flights, and airport size.  

## (1) Data Quality For Tickets and Flights Tables
The following section will overview the quality of the data and some of the issues will be highlighted. 

```{r echo=FALSE, warning=FALSE, message=FALSE}
source("flight_functions.R")
Airports <-
  read_csv("/home/dillon/CodeProjects/R/capitalone/Airport_Codes.csv")
Tickets <-
  read_csv("/home/dillon/CodeProjects/R/capitalone/Tickets.csv")
Flights <-
  read_csv("/home/dillon/CodeProjects/R/capitalone/Flights.csv")
```

- Here is a summary of NA's that exist in the raw data files:

```{r  echo=FALSE, warning=FALSE, message=FALSE}
f <- data.frame(colSums(is.na(Flights)))
f<- add_rownames(f)
f<- rename(f,Name = rowname, NAs = colSums.is.na.Flights..)

g <- data.frame(colSums(is.na(Tickets)))
g<- add_rownames(g)
g<- rename(g,Name = rowname, NAs = colSums.is.na.Tickets..)

kable(f%>%filter(NAs != 0))
kable(g%>% filter(NAs != 0))

```
- NA's for distances of flights. However, this can be overcome by looking through the data. Flights connecting the same two 
airports have good data for distance in the file. Example, BWI to DFW has many NA's, however, the distance is known
from other records in the data:

```{r echo = FALSE}
kable(head(Flights[which(is.na(Flights$DISTANCE)), c("ORIGIN", "DESTINATION", "DISTANCE")]))
kable(head(Flights %>% filter(ORIGIN == "BWI" & DESTINATION == "DFW") %>%
  select(ORIGIN, DESTINATION, DISTANCE)))
```
- The same issue of NA's occurs with ITIN_FARE. Several are missing which
 did cause problems computing an average itinerary fare. These must be omitted. 
- Computing round trips flights is ambiguous with this dataset. Planes that 
return to and from the same airports exist in the data, however, if this 
should qualify as a roundtrip is not clear. In the Tickets dataset passengers
who purchased roundtrip tickets are known by the column ROUNDTRIP, however,
they may have connecting flights in between their origin and destination. 
Therefore, it is difficult to know if flights that are round trip are actually
the best roundtrip routes, since connections are not taken into account. 
The tickets and flights dataset can only be connected through the origin
and roundtrips from the tickets dataset were likely using a multiple flights to reach the destination. 

- There are some records that have impossible air speeds given the flight time and distance. For instance,
 on average one flight was recorded as flying for two minutes but convering a distance of many hundred miles. A function for dealing with this is based on basic physics, $d = rt$. A assumed airspeed of 500 m.p.h is input and a tolerance of 1,000 miles difference on the distance is used.  
 
```{r}
impossible_airspeed <- function(flights){
  c <- 1
  tol <- 1000
  badrows <- list()
  for(i in 1:nrow(flights)){
    d <- 500*(flights$AVG_AIR_TIME[i]/60)
    if(is.na(d)){
      badrows[[c]] <- i 
      c<- c+1
    }
    else{
      diff <- abs(flights$DIST[i] - d)
      if(is.na(diff)){
        badrows[[c]] <- i
        c<-c+1
        break 
      }
      else{
        if( diff > tol ){
          badrows[[c]] <- i
          c<- c+1
        }
      }
    }
  }
  do.call(rbind,badrows)
}
```
 
- Some flights were to and from the same airport. This flight would be useless for the analysis at hand and these records are discarded. 

## (2) Merging the Data
The data is merged together by filtering out origin and destination airports that are too small to fit the criteria. This is done by using the Airports table and merging them on the IATA_CODE. A roundtrip is dealt with by giving a roundtrip a specific code so that either a trip from ORD to LAX is the same code as LAX to ORD. This enables the data to be grouped by this flight ID number and also merged so that the Flights table can be joined to the Tickets table. The data is aggregated over the flight ID number, and averages are taken of Flight Time, Fare, Delay, and other numeric columns. An example of the top rows and first 5 columns is shown next, 

```{r setup, warning=FALSE, message=FALSE, echo=FALSE}

flights <- join_data(Airports, Tickets, Flights)
kable(head(flights[,1:5]))
```

- R and Python suffer from slow computational speeds if function calls are not vectorized. When profiling the code I was able to locate bottle necks in the code at the impute flightid function shown below,

```{r eval=FALSE}
impute_flightid <- function(data, roundtripid, airid_dict) {
  flightid <- numeric(nrow(data))
  if (("ORG_AIRID" %in% colnames(data)) &&
      ("DEST_AIRID" %in% colnames(data))) {
    for (i in 1:nrow(data)) {
      flightid[i] <-
        roundtripid[data$ORG_AIRID[i], data$DEST_AIRID[i]]
    }
    data$flightid <- flightid
    data
  }
  else if (("ORIGIN" %in% colnames(data)) &&
           ("DESTINATION" %in% colnames(data))) {
    for (i in 1:nrow(data)) {
      r <- airid_dict[data$ORIGIN[i], ]
      c <- airid_dict[data$DESTINATION[i], ]
      flightid[i] <- roundtripid[r, c]
    }
    data$flightid <- flightid
    data
  }
  else{
    errorCondition("Necessary columns not included in data")
  }
}
```

- Roughly, the code looks up a trip based on a number given to its origin and destination matrix. A symmetric matix is used to ensure the same trip gets the same id. However, this code is slow because of the use of loops. A typical call of this function takes 7000 ms, on of the slowest function calls in the code. 
- Because of my experience writing C++ code I was able to increase the execution speed of this code using a simple C++ function by 99.9% (10ms on a typical call), 

```{r eval=FALSE}
vector<int> impute_flightid_cpp(DataFrame F, NumericMatrix M, DataFrame Airid) {
  NumericVector Rowid =  F["ORG_AIRID"];
  NumericVector Colid = F["DEST_AIRID"];
  vector<int> flightid; 
  for(int i = 0; i < Rowid.size(); ++i){
    flightid.push_back(M(Rowid(i)-1, Colid(i)-1));
  }
  return flightid; 
}
```

- Because this function needs to be called many times working with this data it is important that the loading of the data happen quickly. The efficiency gains of C++ code within R are substantial and because of libraries like Rcpp, easy to do. 


## (3) Graphs and visualizations
The top ten busiest airports routes based on the number of flights into and out of the same airport are shown: 

```{r echo=FALSE, warning=FALSE, message=FALSE}
top_ten <- flights[1:10,]
top_ten <- cbind(top_ten, unpack_latlon(top_ten))
top_ten$TRIP_NAME <- make_tripname(top_ten)
kable(cbind(Rank = c(1:10),top_ten[,c("N_FLIGHTS", "TRIP_NAME")]))
```

### Visualizing the routes
A graph of the US with the projected routes of the ten busiest flights is shown in the following graph, 

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.align="center"}
library(mapdata)
library(ggrepel)
library(usmap)
usa <- map_data('usa')
ggplot() +
geom_polygon(data = usa,
             aes(x = long, y = lat, group = group),
             fill = 'black') +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid = element_blank(),
    panel.background = element_rect("grey15")
  ) +
  ggtitle('Busiest US Flights') + coord_fixed(1.3) +
  geom_point(data = top_ten,
             aes(x = ORIGIN_LAT, y = ORIGIN_LON),
             color = 'red') + geom_curve(
               curvature = .25,
               data = top_ten,
               aes(
                 x = ORIGIN_LAT,
                 y = ORIGIN_LON,
                 xend = DEST_LAT,
                 yend = DEST_LON
               ),
               col = "orange"
             ) +
  geom_text_repel(data = top_ten,
                  aes(x = ORIGIN_LAT, y = ORIGIN_LON, label = TRIP_NAME),
                  col = "yellow") +
  geom_label_repel(
    data = top_ten,
    aes(x = ORIGIN_LAT, y = ORIGIN_LON, label = N_FLIGHTS),
    arrow = arrow(length = unit(.01, 'npc')),
    box.padding = unit(2.5, 'lines'),
    segment.color = "red"
  )
```

Some issues to note:

- These are not necessarily the most profitable flights. 
- Several records have very few flights or tickets purchased. These are dropped from the analysis since there is not enough data to tell if they are representative. Records with tickets < 75 are removed. I feel there is not enough data on these flights to warrant a strong recommendation.  

### Most profitable flights: 

- The top ten most profitable flights are shown next, 

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.align="center"}
flights$REVENUE <-  compute_flightrev(flights)
flights$COSTS <- compute_flightcost(flights)
flights <-  flights %>% mutate(PROFIT = REVENUE - COSTS)
flights <-  flights %>% mutate(PROFIT = REVENUE - COSTS) %>% filter(N_TICKETS > 75, 
                                                                    PROFIT > 0)
repflights <- flights
repflights$TRIP_NAME <- make_tripname(repflights)
FLIGHTS_TO_BREAKEVEN <- unlist(lapply(repflights$PROFIT, breakeven))
repflights$FLIGHTS_TO_BREAKEVEN <- FLIGHTS_TO_BREAKEVEN
repflights <-
  repflights %>% mutate(SIZE = as.numeric((ORIGIN_SIZE == 'medium_airport' &
                                             DEST_SIZE == 'medium_airport')
  ) +
    2 * as.numeric((
      ORIGIN_SIZE == 'medium_airport' &
        DEST_SIZE == 'large_airport'
    )) +
    3*as.numeric((
      ORIGIN_SIZE == 'large_airport' &
        DEST_SIZE == 'medium_airport'
    )))
repflights$SIZE <- factor(repflights$SIZE)
levels(repflights$SIZE) <- c("Large_to_Large", "Medium_to_Medium", 
                             "Medium_to_Large", "Large_to_Medium")

top_profit <- repflights %>%arrange(desc(PROFIT))
top_profit <- top_profit[1:10, ]
kable(top_profit[,c("PROFIT", "TRIP_NAME", "REVENUE", "COSTS", "FLIGHTS_TO_BREAKEVEN", 
                    "N_FLIGHTS", "DIST", "AVG_ITIN_FARE", "AVG_DEP_DELAY", 
                    "AVG_ARR_DELAY")])
top_profit <- cbind(top_profit, unpack_latlon(top_profit))

top_profit$TRIP_NAME <- make_tripname(top_profit)

ggplot() +
  geom_polygon(data = usa,
               aes(x = long, y = lat, group = group),
               fill = 'black') +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid = element_blank(),
    panel.background = element_rect("grey15")
  ) +
  ggtitle('Most Profitable Trips (Number of Tickets shown in boxes)') + coord_fixed(1.3) +
  geom_point(data = top_profit,
             aes(x = ORIGIN_LAT, y = ORIGIN_LON),
             color = 'red') + geom_curve(
               curvature = .25,
               data = top_profit,
               aes(
                 x = ORIGIN_LAT,
                 y = ORIGIN_LON,
                 xend = DEST_LAT,
                 yend = DEST_LON
               ),
               col = "orange"
             ) +
  geom_text_repel(data = top_profit,
                  aes(x = ORIGIN_LAT, y = ORIGIN_LON, label = TRIP_NAME),
                  col = "yellow") +
  geom_label_repel(
    data = top_profit,
    aes(x = ORIGIN_LAT, y = ORIGIN_LON, label = N_TICKETS),
    arrow = arrow(length = unit(.01, 'npc')),
    box.padding = unit(2.5, 'lines'),
    segment.color = "red"
  )
```

Possible issues:

- Notice that some records show that there are less flights in the dataset than necessary flights to breakeven. This would indicate in one quarter this flight would lose money. In the long run it may be profitable but if short term gains are most important these flights should be avoided. 

### Predictors of profit 

The profit function is defined by $\pi(d, f, r, o_s, d_s, d_d, a_d)$, where $d$ is distance, $f$ is the fare, $r$ is the rate of occupancy, $o_s$ is the origin size, $d_s$ is the destination size, $d_d$ and $a_d$ are the delays in departure and arrival respectively. More specifically, 
$$\pi(d, f, r, o_s, d_s, d_d, a_d) =  r(200f + 7000  ) - 9.44 d - \mathbb{1}_{d_d > 15}\{75(d_d - 15) \} -  \mathbb{1}_{a_d > 15}\{75(a_d - 15) \} - A(o_s) - A(d_s) $$
$$A(x) = \begin{cases}
 10,000 &\text{ if  } x = \text{Large} \\
5,000 & \text{otherwise.}
\end{cases}$$

- Given that this is a linear function in its parameters and estimating supply and demand equations were futile, nonlinear optimization of the profit function is not possible. 
- The most profitable flights will depend linearly on distance, rate, and fare. While airport size is import given that it enters the equation merely as a constant its only effect will be in the way in which it interacts with the other variables. As it will be shown later merely minimizing airport costs is not a viable solution for maximizing profit.  
- Fare depends positively on distance. While distance enters the profit equation as a negative multiplier, fare increases with distance which positively affects profit.
- The positive effect of distance appears to overcome its negative multiplier in profit which is shown below. Some of the most profitable flights are the longer distance flights. 
- The airport size of origin and destination flights is used to break down the data categorically. Consistently the highest profit flights are to and from larger airports. Medium to medium are less profitable. However, when controlling for other factors in a multiple regression the partial effect on profit does reveal an increase in profit (of about 10,000). Yet this is not very surprising given that the partial effect should be around 10,000 dollars as the profit function penalizes large airports by an extra 5,000 on each airport. For this airline, it is my opinion the total of profit is more important than marginal effects. The regression results are shown below, 
```{r warning=FALSE, message=FALSE}
summary(lm(PROFIT ~ N_FLIGHTS + DIST + AVG_ITIN_FARE + SIZE + AVG_ARR_DELAY + AVG_DEP_DELAY, data = repflights))

```


```{r echo=FALSE, warning=FALSE, message=FALSE, fig.width=12}
library(gridExtra)

dfare <- ggplot() + geom_point(data = repflights, aes(x=DIST, y = AVG_ITIN_FARE)) +
  xlab("Distance") + ylab("Average Itinierary Fare") + ggtitle("Fair Paid vs Distance")
dpro <- ggplot() + 
  geom_point(data = repflights, aes(x = DIST, y = PROFIT, colour = SIZE)) +
  scale_colour_manual(values = c("Large_to_Large"="darkgrey", "Large_to_Medium" = "red",
                                 "Medium_to_Medium"="green", "Medium_to_Large"="purple")) +
  ggtitle(label="Profit vs Distance", subtitle = "Colored by Airport Size") +
  geom_smooth(data=repflights, method="lm", aes(x=DIST, y=PROFIT),se=FALSE) +
  geom_jitter() + xlab("Distance") + ylab("Profit")
grid.arrange(dfare, dpro, nrow=1, widths=c(2,2))
```

- This indicates that longer flights are more profitable making it a key characteristic for profitability. However, it is not the most important characteristic. **The most important is high fare price. This difference is summarized in the following two plots:**

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.width=10, fig.height=5}
library(gridExtra)
topflight <-
  repflights %>% mutate(
    TOP_FARES = AVG_ITIN_FARE > quantile(AVG_ITIN_FARE, .90),
    TOP_DIST = DIST > quantile(DIST, .90)
  )
topflight$TOP_DIST <- factor(topflight$TOP_DIST)
levels(topflight$TOP_DIST) <- c("LOW_DISTANCE", "HIGH_DISTANCE")
topflight$TOP_FARES <- factor(topflight$TOP_FARES)
levels(topflight$TOP_FARES) <- c("LOW_FARE", "HIGH_FARE")
p1 <- ggplot() +
  geom_point(data = topflight, aes(x = DIST, y = PROFIT, colour =
                                     TOP_FARES)) + ggtitle("Profit vs Fair Paid") + xlab("Distance") +
  ylab("Profit")+guides(colour=guide_legend(title=">90% In Fares"))

p2 <- ggplot() +
  geom_point(data = topflight, aes(x = AVG_ITIN_FARE, y = PROFIT, colour =
                                     TOP_DIST)) + ggtitle("Profit vs Average Itinerary Fare") +
  xlab("Average Itinerary Fare") + ylab("Profit") + guides(colour=guide_legend(title=">90% In Distance"))

grid.arrange(p1,p2, ncol=2)
```

- The plot on the left shows that profit and distance are positively linearly related. However, when controlling for fare price there are high profit flights that are short and long. This is most clearly shown in the following regression results. When the regression, 
$$\pi = c + \beta d + \varepsilon,$$
where $c$ is a constant, $d$ is distance and $\varepsilon$ is the idiosyncratic error with the usual distributional assumptions the multiplier $\beta$ is positive 5.6. However, when controling for the fare and other factors (included in $X$ and regression results are shown above), 
$$\pi = c + \beta d + \gamma X + \varepsilon,$$
the multiplier on distance is -9.4. Fare is clearly more important than distance. Longer distance flights should only be tolerated if the increase in fare justifies the profit bottom line, in many cases it does. 

- Airport size is another important factor for profit. Strictly looking at the median profit shows that it is increased at larger airports. Roundtrip routes between medium airports did not command high fees in this dataset, although the regression results showed positive marginal effects of smaller airports. Summary plots of these statistics are shown next, 

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.pos="center"}

ggplot(data = repflights, aes(x = SIZE, y = PROFIT, fill = SIZE)) + geom_boxplot() +
  ggtitle("Profit Summary By Airport Size") + xlab("size") + ylab("Profit")
kable(
  repflights %>% group_by(SIZE) %>%
    summarise(
      Min = min(PROFIT),
      Q1 = quantile(PROFIT, .25),
      Median = median(PROFIT),
      Avg = mean(PROFIT),
      Q3 = quantile(PROFIT, .75),
      Max = max(PROFIT) 
    )%>%
    arrange(desc(Median))
)
```

- The most profitable flights in the dataset are longer flights since they come with offsetting high fees for fares. 
  
- Delays obviously have a negative impact on profit. Flights with lower delays will increase profit. However, very profitable flights also have delays, and this should by no means discourage the airline from entering these routes. A investigation into the effect of delays turned up that delays are mainly negligible on profits. The scatter plots show very little impact of delays, and regression results show a statistically significant impact of delays that is not financially meaningful. 
```{r echo=FALSE, warning=FALSE, message=FALSE}
ggplot() +geom_point(data=repflights, aes(x=AVG_ARR_DELAY, y = PROFIT)) + 
  ggtitle("Profit on Arrival Delay")
ggplot() +geom_point(data=repflights, aes(x=AVG_DEP_DELAY, y = PROFIT))+
  ggtitle("Profit on Departure Delay")
```


### Breakeven flights
Breakeven number of flights is another important consideration. Profit is nonlinearly related to this quantity. From the plot of breakeven flights on profit this is clear, 

```{r echo=FALSE, warning=FALSE, message=FALSE}
ggplot() + geom_point(data=topflight, aes(x=FLIGHTS_TO_BREAKEVEN, 
                                          y = PROFIT, colour=TOP_FARES))
```

This function can be estimated by noting that the data transformation of the regression equation allows one to estimate the nonlinear exponent on $x$. Note that,
$$ y = c + \beta x^\gamma + \varepsilon$$
and therefore, 
$$log(y) = c^* + \gamma x + \varepsilon$$
where $c^*$ will be some interaction of $\beta$ and the constant. The regression equation returned that the appropriate $\gamma$ is $-2.02$ and the multiplier is 19.13.
```{r echo=FALSE, warning=FALSE, message=FALSE}
est <- lm(log(PROFIT) ~ log(FLIGHTS_TO_BREAKEVEN), data=repflights)
summary(est)
```

This gives, 
$$ \hat{y} = e^{19.13}\times x^{-2.02}$$
This fits the data quite well. The black line is the fitted equation for the data, 

```{r echo=FALSE, warning=FALSE, message=FALSE}

estfun <- function(fls){
  y <- exp(19.13)*fls^(-2.022)
}
a<- min(topflight$FLIGHTS_TO_BREAKEVEN)
y <- unlist(lapply(a:800,estfun ))
estdata <- data.frame(cbind(X= a:800, Y = unlist(lapply((a:800),estfun ))))

ggplot() + geom_point(data=topflight, aes(x=FLIGHTS_TO_BREAKEVEN, 
                                          y = PROFIT, colour=TOP_FARES))+
  geom_line(data=estdata, aes(x=X, y=Y))
```

This equation is interesting. It shows that there are huge profit gains around the flights that take fewer than 100 flights. In fact, since the derivative of this equation gives, 
$$\frac{dy}{dx} = \frac{-\gamma e^{19.13}}{x^{3.022}}$$
this implies that if it took 50 flight to breakeven a change of just one flight impacts profit by $3,016.80. And since the effect is highly nonlinear the effect impacts profit greatly in the region less than 100 flights but profit is not changed much if the number of flights to break even is much  greater than 100. **This underscores the importance of searching for flights which have a low number of flights to breakeven as the impact on profit is immense by minimizing this quantity.** 

### Recommendations

- The most profitable flights are consistently Large to Large, Large to Medium, Medium to Large airport sizes. Despite there being a positive effect of Medium to Medium airports they cannot command a high enough fare to break even quickly.  
- The most profitable flights are typically longer distance flights. Despite a negative multiplier in the profit function the increase in fares justifies these flights.  
- From this data the number of other available flights does not impact profit. When demand is high the competition does not appear to drive profits to low levels. 
- The most profitable flights will be ones which can command a high itinerary fare. 

Therefore, the recommendation is to track flights which are not medium to medium airport flights, have high fares, are longer distance, and most importantly have a low number of flights to break even. A search of the data for these types of flights came up with the following best five flights for profitability: 

```{r echo=FALSE, warning=FALSE, message=FALSE}
by_profit <- repflights %>% arrange(desc(PROFIT))
kable(by_profit[1:5, c("TRIP_NAME", "ORIGIN_NAME", "DEST_NAME", "PROFIT", "REVENUE", "COSTS", "DIST", "SIZE", "AVG_ITIN_FARE","N_FLIGHTS", "FLIGHTS_TO_BREAKEVEN")])
```
Notice that these flights are all in the top for itinerary fare, distance, go between large airports, and have the fewest number of flights to break even. 

```{r echo=FALSE, warning=FALSE, message=FALSE}
D <- repflights %>%summarise(TOP_PROFIT = quantile(PROFIT, .95),
                        TOP_DIST = quantile(DIST, .95),
                        LOW_BREAKEVEN = quantile(FLIGHTS_TO_BREAKEVEN, .05),
                        TOP_FARE = quantile(AVG_ITIN_FARE, .95))
kable(D)
```

This summary shows the top 95% quantiles for profit, distance, and fare and the lowest 5% quantile for break even flights. All the recommended flights are beyond these levels. 


# Future Research 

The following summary will describe things that should be done in future work to make this research better. 

I attempted many different ways of manipulating the data to estimate supply and demand equations for flights. However, it does not seem like there is an effective way to do this from the data at hand. With an effective equation for demand and supply airports that are undervalued could be identified and exploited. 

Another area that needs further work is introducing nonlinearities in the profit equation for optimization. Profit only depends linearly on distance and other variables. Therefore, nonlinear optimization methods will merely maximize the variables with positive partial effects and shrink to negative infinity the variables with negative partial effects. However, there is clearly a nonlinear effect to these variables. Distance cannot be 0 or even small because then other modes of transportation are viable. Similarly, fare price has a diminishing effect as well. Fare cannot be increased to infinity, people will stop buying and switch to other airlines. Adequate representation of these equations would enable one to optimize the profit in a more effective way. Once this was known flights could be picked that maximized profit and were as close to the maximal values as possible. 

In the future, in order to make sure that these recommendations are the best for the company it will be critical to make sure average itinerary fare is tracked carefully along with occupancy rate. Keeping high occupancy rates above 64% on average and high itinerary fares above $700 per flight would ensure a high payoff even for longer distance flights to high fee airports. 


