---
title: "(4) Recommendations"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: lumen
---
```{r, echo=FALSE, warning=FALSE, message=FALSE}
source("flight_functions.R")

Airports <-
  read_csv("/home/dillon/CodeProjects/R/capitalone/Airport_Codes.csv")
Tickets <-
  read_csv("/home/dillon/CodeProjects/R/capitalone/Tickets.csv")
Flights <-
  read_csv("/home/dillon/CodeProjects/R/capitalone/Flights.csv")
flights <- join_data(Airports, Tickets, Flights)

flights$REVENUE <-  compute_flightrev(flights)
flights$COSTS <- compute_flightcost(flights)
flights <-  flights %>% mutate(PROFIT = REVENUE - COSTS)
flights <-  flights %>% mutate(PROFIT = REVENUE - COSTS) %>% filter(N_TICKETS > 75, 
                                                                    PROFIT > 0)
repflights <- flights
repflights$TRIP_NAME <- make_tripname(repflights)
FLIGHTS_TO_BREAKEVEN <- unlist(lapply(repflights$PROFIT, breakeven))
repflights$FLIGHTS_TO_BREAKEVEN <- FLIGHTS_TO_BREAKEVEN
repflights <-
  repflights %>% mutate(SIZE = as.numeric((ORIGIN_SIZE == 'medium_airport' &
                                             DEST_SIZE == 'medium_airport')
  ) +
    2 * as.numeric((
      ORIGIN_SIZE == 'medium_airport' &
        DEST_SIZE == 'large_airport'
    )) +
    3*as.numeric((
      ORIGIN_SIZE == 'large_airport' &
        DEST_SIZE == 'medium_airport'
    )))
repflights$SIZE <- factor(repflights$SIZE)
levels(repflights$SIZE) <- c("Large_to_Large", "Medium_to_Medium", 
                             "Medium_to_Large", "Large_to_Medium")
```

# Back to main page
- [Link back to main page](capitalone_data.html)

# Recommendations 

- The most profitable flights are consistently Large to Large, Large to Medium, Medium to Large airport sizes. Despite there being a positive effect of Medium to Medium airports they cannot command a high enough fare to break even quickly. 

- The most profitable flights are typically longer distance flights. Despite a negative multiplier in the profit function the increase in fares justifies these flights.  

- From this data the number of other available flights does not impact profit. When demand is high the competition does not appear to drive profits to low levels. 

- The most profitable flights will be ones which can command a high itinerary fare. 

***Therefore, the recommendation is to track flights which are not medium to medium airport flights, have high fares, are longer distance, and most importantly have a low number of flights to break even.*** A search of the data for these types of flights came up with the following best five flights for profitability: 

```{r echo=FALSE, warning=FALSE, message=FALSE}
by_profit <- repflights %>% arrange(desc(PROFIT))
kable(by_profit[1:5, c("TRIP_NAME", "ORIGIN_NAME", "DEST_NAME", "PROFIT", "REVENUE", "COSTS", "DIST", "SIZE", "AVG_ITIN_FARE","N_FLIGHTS", "FLIGHTS_TO_BREAKEVEN")])
```
Notice that these flights are all in the top for itinerary fare, distance, go between large airports, and have the fewest number of flights to break even. 

```{r echo=FALSE, warning=FALSE, message=FALSE}
D <- repflights %>%summarise(TOP_PROFIT = quantile(PROFIT, .95),
                        TOP_DIST = quantile(DIST, .95),
                        LOW_BREAKEVEN = quantile(FLIGHTS_TO_BREAKEVEN, .05),
                        TOP_FARE = quantile(AVG_ITIN_FARE, .95))
kable(D)
```

This summary shows the top 95% quantiles for profit, distance, and fare and the lowest 5% quantile for break even flights. All the recommended flights are beyond these levels. 

# Future Research
- [Link to next step, Future Research](page5_future_research.html)
