---
title: "(2) Merging the Data"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: lumen
---
```{r, echo=FALSE, message=FALSE, warning=FALSE}
source("flight_functions.R")
Airports <-
  read_csv("/home/dillon/CodeProjects/R/capitalone/Airport_Codes.csv")
Tickets <-
  read_csv("/home/dillon/CodeProjects/R/capitalone/Tickets.csv")
Flights <-
  read_csv("/home/dillon/CodeProjects/R/capitalone/Flights.csv")
```
# Back to main page
- [Link back to main page](capitalone_data.html)

# Merging 
- The data is merged together by filtering out origin and destination airports that are too small to fit the criteria. This is done by using the Airports table and merging them on the IATA_CODE. A roundtrip is dealt with by giving a roundtrip a specific code so that either a trip from ORD to LAX is the same code as LAX to ORD. This enables the data to be grouped by this flight ID number and also merged so that the Flights table can be joined to the Tickets table. The data is aggregated over the flight ID number, and averages are taken of Flight Time, Fare, Delay, and other numeric columns. An example of the top rows and first 5 columns is shown next, 

- The join_data function was built keeping the "builder mindset" goal in mind. It is a function that takes Airports, Tickets, and Flights tables and merges them all together based on characteristics needed for analysis. It is a function that calls many different subroutines. 
```{r}
join_data <- function(AirportData, TicketData, FlightData) {
  big_airports <- clean_airports(AirportData)
  valid_flights <- clean_flights(FlightData, big_airports)
  tickets_roundtrip <- clean_tickets(TicketData, big_airports)
  uniquelevels <- unique(c(
    levels(factor(valid_flights$ORIGIN)),
    levels(factor(valid_flights$DESTINATION)),
    levels(factor(tickets_roundtrip$ORIGIN)),
    levels(factor(tickets_roundtrip$DESTINATION))
  ))
  airid_dict <- make_airid_dict(uniquelevels)
  airid <- airid_dict %>% mutate(NAME = rownames(airid_dict))
  rownames(airid) <- 1:nrow(airid)
  roundtripidmatrix <- make_roundtripid_matrix(uniquelevels)
  roundtripmatrix <<- roundtripidmatrix
  airid <<- airid
  valid_flights <- connect_airid(valid_flights, airid)
  fid <- impute_flightid_cpp(valid_flights, roundtripidmatrix)
  valid_flights$flightid <- fid
  # valid_flights <- impute_flightid(valid_flights, roundtripidmatrix, airid)
  tickets_roundtrip <- connect_airid(tickets_roundtrip, airid)
  # tickets_roundtrip <- impute_flightid(tickets_roundtrip, roundtripidmatrix, airid)
  fid <- impute_flightid_cpp(tickets_roundtrip, roundtripidmatrix)
  tickets_roundtrip$flightid <- fid
  # Now the tables are connected through the airid
  valid_flights <-
    valid_flights %>% mutate(
      DEP_DELAY = as.numeric(DEP_DELAY),
      ARR_DELAY = as.numeric(ARR_DELAY),
      OCCUPANCY_RATE = as.numeric(OCCUPANCY_RATE),
      DISTANCE = as.numeric(DISTANCE),
      AIR_TIME = as.numeric(AIR_TIME),
      DEST_SIZE = factor(DEST_SIZE),
      ORIGIN_SIZE = factor(ORIGIN_SIZE)
    )
  # Aggregate data down to link together
  busy_route <- valid_flights %>% group_by(ORIGIN, DESTINATION) %>%
    summarise(
      N_FLIGHTS = n(),
      ORIGIN_SIZE = mode(ORIGIN_SIZE),
      DEST_SIZE = mode(DEST_SIZE),
      flightid = mode(flightid)
    )
    summarydata <-
    valid_flights %>% group_by(flightid) %>%
    summarise(
      AVG_DEP_DELAY = mean(DEP_DELAY),
      AVG_ARR_DELAY = mean(ARR_DELAY, na.rm = TRUE),
      AVG_AIR_TIME = mean(AIR_TIME, na.rm = TRUE),
      DIST = mode(DISTANCE),
      AVG_OCCUPANCY = mean(OCCUPANCY_RATE, na.rm = TRUE),
    )
  
  flights_roundtrips <- busy_route %>% group_by(flightid) %>%
    summarise(
      N_FLIGHTS = min(N_FLIGHTS),
      ORIGIN = mode(ORIGIN),
      DESTINATION = mode(DESTINATION)
    ) %>%
    arrange(desc(N_FLIGHTS))
  flights_roundtrips <- flights_roundtrips %>%
    inner_join(big_airports[, c(1, 2, 4, 5)], by = c("ORIGIN" = "IATA_CODE")) %>%
    rename(
      "ORIGIN_NAME" = "NAME",
      "ORIGIN_SIZE" = "TYPE",
      "ORIGIN_COOR" = "COORDINATES"
    ) %>%
    inner_join(big_airports[, c(1, 2, 4, 5)], by = c("DESTINATION" = "IATA_CODE")) %>%
    rename(
      "DEST_NAME" = "NAME",
      "DEST_SIZE" = "TYPE",
      "DEST_COOR" = "COORDINATES"
    )
  
  tickets_roundtrip <-
    tickets_roundtrip[which(!is.na(tickets_roundtrip$ITIN_FARE)), ]
  ticket_summary <- tickets_roundtrip %>% group_by(flightid) %>%
    summarise(N_TICKETS = n(),
              AVG_ITIN_FARE = mean(ITIN_FARE)) %>%
    arrange(desc(N_TICKETS))
  
  flights_roundtrips <-
    inner_join(flights_roundtrips,
               summarydata,
               by = c("flightid" = "flightid"))
  ## Actually connects the three tables together
  flights_roundtrips <- flights_roundtrips %>% 
    inner_join(ticket_summary, by=c("flightid" = "flightid"))
  
  flights_roundtrips <- flights_roundtrips %>% select(
    flightid,
    N_FLIGHTS,
    N_TICKETS,
    ORIGIN,
    DESTINATION,
    ORIGIN_SIZE,
    DEST_SIZE,
    ORIGIN_COOR,
    DEST_COOR,
    ORIGIN_NAME,
    DEST_NAME,
    AVG_ITIN_FARE,
    AVG_DEP_DELAY,
    AVG_ARR_DELAY,
    AVG_AIR_TIME,
    AVG_OCCUPANCY,
    DIST
  )
  bad <- impossible_airspeed(flights_roundtrips)
  flights_roundtrips <- flights_roundtrips[-bad, ]
  
  flights_roundtrips
}

```

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
flights <- join_data(Airports, Tickets, Flights)
kable(head(flights[,1:5]))
```


- To see [Functions realted to AIRID Click Here.](code_documentation.html)

# Graphs and Visualizations
- [Link to step 3 Graphs and Visualizations](page3_graphs.html)






