---
title: "(1) Data Quality For Tickets and Flights Tables"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: lumen
---
# Back to main page
- [Link back to main page](capitalone_data.html)

# Import 
- The following section will overview the quality of the data and some of the issues will be highlighted. 

```{r  warning=FALSE, message=FALSE}
source("flight_functions.R")
Airports <-
  read_csv("/home/dillon/CodeProjects/R/capitalone/Airport_Codes.csv")
Tickets <-
  read_csv("/home/dillon/CodeProjects/R/capitalone/Tickets.csv")
Flights <-
  read_csv("/home/dillon/CodeProjects/R/capitalone/Flights.csv")
```
# NA's
- Here is a summary of NA's that exist in the raw data files:

```{r echo=FALSE, warning=FALSE, message=FALSE}
f <- data.frame(colSums(is.na(Flights)))
f<- add_rownames(f)
f<- rename(f,Name = rowname, NAs = colSums.is.na.Flights..)

g <- data.frame(colSums(is.na(Tickets)))
g<- add_rownames(g)
g<- rename(g,Name = rowname, NAs = colSums.is.na.Tickets..)

kable(f%>%filter(NAs != 0))
kable(g%>% filter(NAs != 0))

```
- NA's for distances of flights. However, this can be overcome by looking through the data. Flights connecting the same two 
airports have good data for distance in the file. Example, BWI to DFW has many NA's, however, the distance is known
from other records in the data:

```{r echo = FALSE}
kable(head(Flights[which(is.na(Flights$DISTANCE)), c("ORIGIN", "DESTINATION", "DISTANCE")]))
kable(head(Flights %>% filter(ORIGIN == "BWI" & DESTINATION == "DFW") %>%
  select(ORIGIN, DESTINATION, DISTANCE)))
```
- The same issue of NA's occurs with ITIN_FARE. Several are missing which
did cause problems computing an average itinerary fare. These must be omitted.

# Roundtrips calculation
- Computing round trips flights is ambiguous with this dataset. Planes that 
return to and from the same airports exist in the data, however, if this 
should qualify as a roundtrip is not clear. In the Tickets dataset passengers
who purchased roundtrip tickets are known by the column ROUNDTRIP, however,
they may have connecting flights in between their origin and destination. 
Therefore, it is difficult to know if flights that are round trip are actually
the best roundtrip routes, since connections are not taken into account. 
The tickets and flights dataset can only be connected through the origin
and roundtrips from the tickets dataset were likely using a multiple flights to reach the destination. 

# Impossible Airspeeds 
- There are some records that have impossible air speeds given the flight time and distance. For instance,
on average one flight was recorded as flying for two minutes but convering a distance of many hundred miles. A function for dealing with this is based on basic physics, $d = rt$. A assumed airspeed of 500 m.p.h is input and a tolerance of 1,000 miles difference on the distance is used.  

```{r}
impossible_airspeed <- function(flights){
  c <- 1
  tol <- 1000
  badrows <- list()
  for(i in 1:nrow(flights)){
    d <- 500*(flights$AVG_AIR_TIME[i]/60)
    if(is.na(d)){
      badrows[[c]] <- i 
      c<- c+1
    }
    else{
      diff <- abs(flights$DIST[i] - d)
      if(is.na(diff)){
        badrows[[c]] <- i
        c<-c+1
        break 
      }
      else{
        if( diff > tol ){
          badrows[[c]] <- i
          c<- c+1
        }
      }
    }
  }
  do.call(rbind,badrows)
}
```

# Other Issues 
- Some flights were to and from the same airport. This flight would be useless for the analysis at hand and these records are discarded. 

# Next steps - Merging
- [Link step 2 of the project, Merging](page2_merging.html)